GEE models

### 2016-06-29 doc/manuscript.Rmd
Updated manuscript to reflect DC reqs plus Tony's edits

Word count: 3543 / 4000

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

@@

sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. GEE models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

@@

ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')

@@

# Subjects and Methods

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
While full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
continued to show negative associations with IGI/IR and ISSI-2. Sensitivity
analyses revealed that adjustment for WC was responsible for attenuating the
unadjusted results. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD increase
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. However, in contrast to these results using NEFA modeled as
concentrations, none of the NEFA variables modeled as mol% were associated with
the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
ISSI-2 (data not shown). Sensitivity analyses to examine the association with
the sum of fatty acid classes (saturated vs unsaturated) revealed similar
findings as seen with the adjusted GEE model for total NEFA (data not shown).
There were no significant interaction effects between time, sex, or ethnicity
and the individual NEFA on any of the outcome measures, except for a single
marginally significant (p=0.04) interaction between sex, 20:3n-6 (mol%), and
ISI, which showed that males had a 12% higher ISI for every SD increase in
20:3n-6. Raw values from the GEE models are shown in `r cite_st('gee')` for
fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis, which was used to confirm the results
of the GEE modeling, but in a multivariate environment (all NEFA in a single
model). The PLS-DA results show a very poor discriminatory ability of the
NEFA composition in classifying participants into the correct beta-cell function
latent class. The percent of participants correctly predicted to belonging to
their group (i.e. those predicted in the low group and actually belonging to the
low group) was `r misclass.n.percent.bcf$Percent[9]` in the low group,
`r misclass.n.percent.bcf$Percent[5]` in the middle group, and 
`r misclass.n.percent.bcf$Percent[1]` in the high group (the distribution of the
groupings based on the components from PLS-DA are shown in `r cite_sf('plsda_groups')`). 
The loadings plot of the NEFA composition against the extracted PLS-DA
components is shown in `r cite_sf('plsda_loadings')` and is used to identify
fatty acids that contribute the most explained variance to the beta-cell
function groupings. However, because of the high rate of misclassification, no
meaningful predictive information can be extracted even if any NEFA does
contribute to the explained variance (e.g. 16:0). The PLS-DA analysis using
conversion to or maintenance of dysglycemia status over the 6 years as the
response variable showed similar poor discriminatory ability, though better than
the beta-cell function trajectory results. While 
`r misclass.n.percent.dysgly$Percent[1]`
of participants were correctly predicted to not have dysglycemia, only 
`r misclass.n.percent.dysgly$Percent[4]` were correctly predicted to have
dysglycemia (see `r cite_sf('plsda_groups_dysgly')` for the distribution of 
dysglycemia status based on the components from the PLS-DA and 
`r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
by extracted PLS-DA components).

# Discussion

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA concentrations independently predicted
lower beta-cell function after 6 years. While we found negative associations with
palmitic acid (16:0), oleic acid (18:1n-9), and
linoleic acid (18:2n-6) for the concentration (nmol/mL) modeling, no
associations were seen for the proportion of these fatty acids and further
clustering analysis found no predictive ability of these individual NEFA on
beta-cell function or dysglycemia. These observations suggest that the absolute
size of the total NEFA fraction, rather than its specific composition, likely
influences the pathogenesis of diabetes.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified that
adjustment for WC was likely the cause for the attenuation in results between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, there may be differences in physiology between fasting and 
postprandial NEFA kinetics that we are not able to investigate but that may 
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Second, the
null findings may be due to the high risk population examined in PROMISE,
including having a higher WC. It may be that in this population, IR has become
well established and NEFA may not contribute to IR at this somewhat more
advanced stage in the pathogenesis of diabetes.

@@

The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
in Family Medicine, University of Western Ontario, London, Canada for their 
expert technical assistance and dedication in their work for PROMISE. 
The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript.  The
authors report no potential conflicts of interest relevant to this study.

This study was supported by grants from the Canadian Diabetes Association (CDA),
the Canadian Institutes for Health Research (CIHR), and the University of
Toronto Banting and Best Diabetes Centre (BBDC); LWJ is supported by a CDA
Doctoral Student Research Award; RR is supported by a Heart and Stroke
Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
National Diabetes Management and the Ian McWhinney Chair of Family Medicine
Studies at the University of Western Ontario; RBP holds a Tier II Canada
Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
Chair in Diabetes Epidemiology.

@@

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results)
```

@@

## Supplemental Material

@@

The two plots generated from the PLS-DA analysis show the results from both
the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
the predictors (all 22 NEFA). The response result plots as seen in 
`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
discriminatory ability of the PLS-DA model. When the predictors represent an
underlying pattern that may exist between groups, the plots will show greater
separation between the groups in addition to a low misclassification rate. The
predictor result plots as seen in `r cite_sf('plsda_loadings')` and
`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
contribute the most to the explained variance. Variables between the solid and 
dashed lines explain between 50-100% of the explained variance. Whether the 
variables have positive or negative values on either of the components axes 
determines how they contribute to each group. For instance, if those individuals
with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
has a positive value for component 1, this indicates that 16:0 is predictive of
having dysglycemia.

## Supplemental Data

@@

```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings}
plot_plsda_loadings(plsda_results)
```

@@

```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly}
plot_plsda_loadings(plsda_dysgly)
```

>>>>

Word count: 3602 / 4000

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')

@@

sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. GEE models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association (*).')

@@

ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk (*).')

@@

# Research Design and Methods

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
While full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
continued to show negative associations with IGI/IR and ISSI-2. Sensitivity
analyses revealed that adjustment for WC was responsible for attenuating the
unadjusted results. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD increase
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. However, in contrast to these results using NEFA modeled as
concentrations, none of the NEFA variables modeled as mol% were associated with
the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
ISSI-2 (data not shown). Sensitivity analyses to examine the association with
the sum of fatty acid classes (saturated vs unsaturated) revealed similar
findings as seen with the adjusted GEE model for total NEFA (data not shown).
There were no significant interaction effects between time, sex, or ethnicity
and the individual NEFA on any of the outcome measures, except for a single
marginally significant (p=0.04) interaction between sex, 20:3n-6 (mol%), and
ISI, which showed that males had a 12% higher ISI for every SD increase in
20:3n-6 compared to females. Raw values from the GEE models are shown in `r cite_st('gee')` for
fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of these groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis, which was used to confirm the results
of the GEE modeling, but in a multivariate environment (all NEFA in a single
model). The PLS-DA results show a very poor discriminatory ability of the
NEFA composition in classifying participants into the correct beta-cell function
latent class. The percent of participants correctly predicted to belonging to
their group (i.e. those predicted in the low group and actually belonging to the
low group) was `r misclass.n.percent.bcf$Percent[9]` in the low group,
`r misclass.n.percent.bcf$Percent[5]` in the middle group, and 
`r misclass.n.percent.bcf$Percent[1]` in the high group (the distribution of the
groupings based on the components from PLS-DA are shown in `r cite_sf('plsda_groups')`). 
The loadings plot of the NEFA composition against the extracted PLS-DA
components is shown in `r cite_sf('plsda_loadings')` and is used to identify
fatty acids that contribute the most explained variance to the beta-cell
function groupings. However, because of the high rate of misclassification, no
meaningful predictive information can be extracted even if any NEFA does
contribute to the explained variance (e.g. 16:0). The PLS-DA analysis using
conversion to or maintenance of dysglycemia status over the 6 years as the
response variable showed similar poor discriminatory ability, though better than
the beta-cell function trajectory results. While 
`r misclass.n.percent.dysgly$Percent[1]`
of participants were correctly predicted to not have dysglycemia, only 
`r misclass.n.percent.dysgly$Percent[4]` were correctly predicted to have
dysglycemia (see `r cite_sf('plsda_groups_dysgly')` for the distribution of 
dysglycemia status based on the components from the PLS-DA and 
`r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
by extracted PLS-DA components). The results from the PLS-DA analysis are consistent
with what the associations seen in the GEE model.

# Conclusions

In a Canadian population of adults who are at-risk for diabetes, we found that 
higher total NEFA concentrations independently predicted lower beta-cell
function after 6 years. While we found negative associations with palmitic acid
(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
(nmol/mL) modeling, no associations were seen for the proportion of these fatty
acids and further confirmed by clustering analysis found no predictive ability of these
individual NEFA on beta-cell function or dysglycemia. These observations suggest
that the absolute size of the total NEFA fraction, rather than its specific
composition, likely influences the pathogenesis of diabetes, at least within a
population at-risk for diabetes.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw no association of any individual or total NEFA with hepatic
(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified
WC as having a powerful attenuating effect on the associations between
unadjusted and adjusted models. There are some possible explanations for our
null findings for the insulin sensitivity measures and the influence of WC on
the results. First, WC may be a strong causal link in previously reported
associations between NEFA and insulin sensitivity, given the role adipocytes
play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
in physiology between fasting and postprandial NEFA kinetics that we are not
able to investigate but that may explain our null findings for fasting NEFA and
insulin sensitivity. For instance, some experimental studies using clamp
protocols found that fasting NEFA was a weak predictor of insulin sensitivity
compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
metabolically active given postprandial activity. Third, the null findings seen
may be due to the high risk population examined in PROMISE, including having a 
high WC and BMI. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat more advanced
stage in the pathogenesis of diabetes.

@@

The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
in Family Medicine, University of Western Ontario, London, Canada for their 
expert technical assistance and dedication in their work for PROMISE. 
The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, ZL, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript. AJH is
the guarantor of this work and, as such, had full access to all the data in the
study and takes responsibility for the integrity of the data and the accuracy of
the data analysis.

This study was supported by grants from the Canadian Diabetes Association (CDA),
the Canadian Institutes for Health Research (CIHR), and the University of
Toronto Banting and Best Diabetes Centre (BBDC); LWJ is supported by a CDA
Doctoral Student Research Award; RR is supported by a Heart and Stroke
Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
National Diabetes Management and the Ian McWhinney Chair of Family Medicine
Studies at the University of Western Ontario; RBP holds a Tier II Canada
Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
Chair in Diabetes Epidemiology. The authors report no potential conflicts of
interest relevant to this study.

Portions of the results in the present analysis have been presented in
abstract form at the following conferences: International Symposium on Diabetes
& Nutrition (2015); Biennial Congress of the International Society for the Study
of Fatty Acids and Lipids (2014); International Diabetes Federation World
Diabetes Congress (2015); American Diabetes Association Scientific Sessions
(2014, 2015).

@@

```{r figGEE, dependson='geeData', fig.cap=fig_gee}
plot_gee_results(gee_results, leg = 'none')
```

@@

## Supplemental Methods

@@

The two plots generated from the PLS-DA analysis show the results from both
the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
the predictors (all 22 NEFA). The response result plots as seen in 
`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
showing an example of good discriminatory ability. When the predictors represent an
underlying pattern that may exist between groups, the plots will show greater
separation between the groups in addition to a low misclassification rate. The
predictor result plots as seen in `r cite_sf('plsda_loadings')` and
`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
contribute the most to the explained variance. Variables between the solid and 
dashed lines explain between 50-100% of the explained variance. Whether the 
variables have positive or negative values on either of the components axes 
determines how they contribute to each group. For instance, if those individuals
with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
has a positive value for component 1, this indicates that 16:0 is predictive of
having dysglycemia.

## Supplemental Tables and Figures

@@

```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_results)
```

@@

```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
plot_plsda_loadings(plsda_dysgly)
```
